---
Title: MATH5340 Assessment
Author: R 
Description: R Markdown Project file for MATH5340 Assessment
---

# Library Calls 

Required libraries for this project. 

```{r}
## tidyverse has good dml through dplyr and lubridate
## and dataviz through ggplot2
suppressMessages(library(tidyverse))
suppressMessages(library (here)) # here just enables reliable pathfinding
suppressMessages(library(MASS))  # MASS provides multivariate Gaussian simulations

```

# Question 1

We'll have to assess VaR and AVaR, both at the 99% percentile. We've got 4 stocks, denoted \lambda's. We are given closing prices and shares below:


```{r}
close_prices <- read_csv(here("data/Data_Student_201605543.csv"))

head(close_prices)

shares <- read_csv(here("data//share_numbers.csv"))

my_shares <- shares %>% filter(`Student ID` == 201605543)
```

Now we want to make a matrix of our risk factor changes over time. We'll do this by creating a logged version of the data and taking differences.

```{r}
pclose <- log(close_prices[,-1])
print(pclose)

dates <- close_prices$Date[1:(length(close_prices$Date)-1)]

risk_factors = lapply(pclose, diff)
risk_factors = tibble( # I just like tibbles
    `Date`    = dates,
    `Asset_1` = risk_factors$"Asset 1",
    `Asset_2` = risk_factors$"Asset 2",
    `Asset_3` = risk_factors$"Asset 3",
    `Asset_4` = risk_factors$"Asset 4"
)

risk_factors
```

## (i)

We are now tasked with cleaning the dataset from obvious outliers. I would imagine the first step is to visualise the data. We should start with scatter and violin plots for the dataset and move from there.

```{r, echo = FALSE}
## Seeing 6 clear outliers in Asset 1
ggplot(risk_factors, aes(x = Date, y = Asset_1)) +
    geom_point()

## Seeing what I believe to be the same 6 outliers
ggplot(risk_factors, aes(x = Date, y = Asset_2)) +
    geom_point()

## third verse same as the first
ggplot(risk_factors, aes(x = Date, y = Asset_3)) +
    geom_point()

## I wonder if he simulated Brownian motion processes
## and manually made specific outliers for each of us
ggplot(risk_factors, aes(x = Date, y = Asset_4)) +
    geom_point()
```

All of our outliers seem to have the same property of |x| > 1 where `x` is a risk factor. I propose we simply dplyr filter for this property.

```{r, echo = FALSE}
risk_factors <- risk_factors %>% dplyr::filter(Asset_1 < 1) %>%
    dplyr::filter(Asset_1 > -1) # these are all the same outliers

ggplot(risk_factors, aes(x = Date, y = Asset_1)) +
    geom_point()

ggplot(risk_factors, aes(x = Date, y = Asset_2)) +
    geom_point()

ggplot(risk_factors, aes(x = Date, y = Asset_3)) +
    geom_point()

ggplot(risk_factors, aes(x = Date, y = Asset_4)) +
    geom_point()

## These all look good now
```

We notice more volatility in 2020 but that's to be expected really. Interesting risk plateau in mid-2021, was that the end of the second lockdown or just a simulation artefact?

## (ii)

We are now tasked with applying the variance-covariance method for estimating VaR and AVaR. We are given the date-range "21/02/2021-21/02/2023". We assume that the risk factors are i.i.d.. 


The first thing we want to do is load a variable with the right date range.

```{r, echo = FALSE}
start_date_1 <- dmy("21-02-2021")
end_date_1 <- dmy("21-02-2023")

start_date_2 <- dmy("10-06-2019")
end_date_2 <- dmy("10-06-2021")

load_dates_1 <- risk_factors %>% dplyr::filter(Date > start_date_1) %>%
    dplyr::filter(Date < end_date_1) 

load_dates_2 <- risk_factors %>% dplyr::filter(Date > start_date_2) %>%
    dplyr::filter(Date < end_date_2) 

head(load_dates_1)
tail(load_dates_1) # both final dates not predicted day
head(load_dates_2)
tail(load_dates_2)
```

We want to apply the variance-covariance method to `load_dates` now. For this, we have to assume normal stock returns and linearised profit, big assumptions. We begin by obtaining a mean vector and an estimated covariance matrix for the dataset. We also want the prices on the last day of the loaded date matrix.

```{r, echo = FALSE}
## 21/02/2023
load_means <- colMeans(load_dates_1[sapply (load_dates_1, is.numeric)]) # our means vector 

load_covs <- cov(load_dates_1[-1]) # our covariance matrix

final_date <- as.data.frame(load_dates_1)[dim(load_dates_1)[1], 1]
final_close <- close_prices %>% dplyr::filter(Date == final_date)
final_close # close prices for the day before our predicted day

my_shares # my share prices in every asset

alpha <- 0.01 # working with 99% CI 
```

We can now apply the variance-covariance method to the first date.

```{r, echo = FALSE} 
lin_mu <- as.numeric(my_shares[,-1] * final_close[,-1]) %*% as.vector(load_means)
lin_mu

lin_sigma <- sqrt((as.numeric(my_shares[,-1] * final_close[,-1])) %*%
as.matrix(load_covs) %*% (as.numeric(my_shares[,-1] * final_close[,-1])))
lin_sigma

lin_VaR <- -lin_mu + lin_sigma * qnorm (1-alpha)
lin_VaR

lin_AVaR <- -lin_mu + lin_sigma * dnorm(qnorm (1-alpha))/alpha
lin_AVaR

```

We can just repeat this for the second set of dates and obtain an objective solution.


### (a)

Firstly, we should fit marginal t-distributions to each of our stocks. MASS supplies a fitdistr function that uses MLE techniques to fit known distributions (including the Pearson family).

```{r, echo = FALSE}
set.seed(1234)

fitting_marginals <- function (series) {
    ## just a set of test dfs
    trial_dfs <- c(2, 4, 10, 16, 32, 64) 
    df_log_lik <- c(1:6)

    for (i in c(1:6)) {
        ## MASS t distribution fitting
        fit <- fitdistr(
            series,
            "t",
            df = trial_dfs[i], # optional df
            start = list(
                m  = median(series),
                s  = IQR(series) / 2
            ),
            lower = c(m = -Inf, s = 1e-8)
        )

        df_log_lik[i] = fit$loglik
    }
    
    return (trial_dfs[which.max(df_log_lik)])
}

## all of these came out as 4 df being the best for me
fitting_marginals(load_dates_1$Asset_1)
fitting_marginals(load_dates_1$Asset_2)
fitting_marginals(load_dates_1$Asset_3)
fitting_marginals(load_dates_1$Asset_4)

fit_1 <- fitdistr (load_dates_1$Asset_1, "t", df = 4)
fit_1$estimate

fit_2 <- fitdistr (load_dates_1$Asset_2, "t", df = 4,
                   start = list(
                       m  = median(load_dates_1$Asset_2),
                       s  = IQR(load_dates_1$Asset_2) / 2),
                   lower = c(m = -Inf, s = 1e-8))
fit_2$estimate

fit_3 <- fitdistr (load_dates_1$Asset_3, "t", df = 4)
fit_3$estimate

fit_4 <- fitdistr (load_dates_1$Asset_4, "t", df = 4,
                   start = list(
                       m  = median(load_dates_1$Asset_4),
                       s  = IQR(load_dates_1$Asset_4) / 2),
                   lower = c(m = -Inf, s = 1e-8))
fit_4$estimate


```

Play it safe with the Gaussian. Looking at the distribution of our asset returns:

```{r, echo = FALSE}
hist(risk_factors$Asset_1, breaks=100)
```

Assuming _t_-distributed marginals seems fair. This gives us a known quantile function qt() in R. 

```{r, echo = FALSE}
simulations = 10000
copula_rho <- 0.7 # will be estimated later
copula_R <- matrix (c(1, copula_rho, copula_rho, copula_rho, copula_rho, 1, copula_rho, copula_rho, copula_rho, copula_rho, 1, copula_rho, copula_rho, copula_rho, copula_rho, 1), 4, 4) # first time btw
print(copula_R)

stock1_v <- 4
stock2_v <- 4
stock3_v <- 4
stock4_v <- 4

cat("Determinants:")
print(c(det(copula_R), det(copula_R[1:3, 1:3]), det(copula_R[1:2, 1:2]), 1))

## Copula sampling distribution
z_samples <- MASS::mvrnorm(n = simulations, mu = rep(0, 4),
                           Sigma = copula_R)

## chisq_samples <- rchisq(n = simulations, df = copula_v)

## t_samples = sweep(z_samples, 1,
##                   sqrt(chisq_samples / copula_v), "/") # sweep runs static matrix operations

## These are our marginal distributions
par(mfrow = c(4,1))
hist(z_samples[,1], breaks = 100)
hist(z_samples[,2], breaks = 100)
hist(z_samples[,3], breaks = 100)
hist(z_samples[,4], breaks = 100)
par(mfrow = c(1,1))

## These are our uniform distributions
u <- pnorm(z_samples)
par(mfrow = c(4,1))
hist(u[,1], breaks = 100)
hist(u[,2], breaks = 100)
hist(u[,3], breaks = 100)
hist(u[,4], breaks = 100)
par(mfrow = c(1,1))

## These are our (prospective) modelled stocks
stock1_dist <- fit_1$estimate[1] +
    fit_1$estimate[2] * qt(u[,1], df = stock1_v)
stock2_dist <- fit_2$estimate[1] +
    fit_2$estimate[2] * qt(u[,1], df = stock2_v)
stock3_dist <- fit_3$estimate[1] +
    fit_3$estimate[2] * qt(u[,1], df = stock3_v)
stock4_dist <- fit_4$estimate[1] +
    fit_4$estimate[2] * qt(u[,1], df = stock4_v)


par(mfrow = c(4,1))
hist(stock1_dist, breaks = 100)
hist(stock2_dist, breaks = 100)
hist(stock3_dist, breaks = 100)
hist(stock4_dist, breaks = 100)
par(mfrow = c(1,1))

## the real flexibility of the meta t-distribution is
## free dofs

```

In Lecture 7, Dr. Ararat presented the uniform marginal distributions on a scatterplot to demonstrate the relation.

```{r, echo = FALSE} 

par(mfrow = c(2,2))
plot(u[,1], u[,2], pch = '.')
plot(u[,2], u[,3], pch = '.')
plot(u[,3], u[,4], pch = '.')
plot(u[,1], u[,4], pch = '.')
par(mfrow = c(1,1))

pairs(u, pch = ".", main = "Scatterplot Matrix of Uniform Variables")

```

We can also show the marginal distributions (Lec 8).

```{r, echo = FALSE} 

par(mfrow = c(2,2))
plot(stock1_dist, stock2_dist, pch = '.')
plot(stock2_dist, stock3_dist, pch = '.')
plot(stock3_dist, stock4_dist, pch = '.')
plot(stock1_dist, stock4_dist, pch = '.')
par(mfrow = c(1,1))

```

They look good but there is an inflated concordance since all of the degrees of freedom are currently set to 4.



## (v)

### (b)

On page 176 they define the fancy rho function as:

Y_{t, j} = \Phi^{-1} (\hat{U}_{t,j}) for j in dimensions

\det is the square rooted trace

rho (\sum) = (\det (\sigma))^{-1} \sigma (\det (\sigma))^{-1}

On page 272 they show that the MLE estimator of the correlation matrix is 

\hat{\sum} = n^{-1} \sum^n_{t=1} Y_t Y_t'

With a better approximation being rho (\sum)

We have a relatively simple roadmap for a function based on McNeil et al.
1. We want to fit t-distributions to each return,
2. We want to convert the data to uniform distributions across the marginals,
3. We want to perform a d-dimensional Gaussian quantile transformation into the Copula space,
4. We want to calculate the crossproduct (covariance) of the Gaussian quantile distribution,
5. We want to calculate our determinant of our experimental covariance as the trace of the square rooted matrix,
6. We want to return the correlation as 

```{r, echo = FALSE}
U_1 <- pt((load_dates_1$Asset_1 - fit_1$estimate[1])
          / fit_1$estimate[2], df = 4)
U_2 <- pt((load_dates_1$Asset_2 - fit_2$estimate[1])
          / fit_2$estimate[2], df = 4)
U_3 <- pt((load_dates_1$Asset_3 - fit_3$estimate[1])
          / fit_3$estimate[2], df = 4)
U_4 <- pt((load_dates_1$Asset_4 - fit_4$estimate[1])
          / fit_4$estimate[2], df = 4)

U <- cbind(U_1, U_2, U_3, U_4)
Y <- qnorm(U)

sigma_hat <- crossprod(Y) / nrow(Y)

Det <- diag(1 / sqrt(diag(sigma_hat)))
P_hat <- Det %*% sigma_hat %*% Det
print("Our estimated correlation matrix: ")
print(P_hat)

c_means <- rep(0,4)
print("Our copula means: ")
print(c_means)

```

We now have a Gaussian copula (by default centred on 0) with a fitted correlation matrix. Our marginal t-distributions also have estimated parameters, allowing for free conversion between the uniform and distribution functions.

The obvious path forwards now is to generate a set of uniforms from the copula and convert those back into the marginal distributions, estimate the total portfolio value change, sort them, and select the one at the (1-alpha) proportion.

```{r, echo = FALSE} 
simulations <- 10000
alpha <- 0.99
day_value <- close_prices %>% dplyr::filter (Date == end_date_1)

## take a joint estimate
joint_estimate <- MASS::mvrnorm(n = simulations, mu = c_means, Sigma = P_hat)

## convert to the marginal distributions
hist(qt(pnorm(joint_estimate[,1]), df = 4), breaks=100) 

marginals_1 <- qt(pnorm(joint_estimate[,1]), df = 4) * fit_1$estimate[2] + fit_1$estimate[1]
marginals_2 <- qt(pnorm(joint_estimate[,2]), df = 4) * fit_2$estimate[2] + fit_2$estimate[1]
marginals_3 <- qt(pnorm(joint_estimate[,3]), df = 4) * fit_3$estimate[2] + fit_3$estimate[1]
marginals_4 <- qt(pnorm(joint_estimate[,4]), df = 4) * fit_4$estimate[2] + fit_4$estimate[1]

estimated_returns <- cbind(marginals_1, marginals_2, marginals_3, marginals_4)

head(estimated_returns)
par(mfrow = c(4, 1))
hist(estimated_returns[,1], breaks = 100)
hist(estimated_returns[,2], breaks = 100)
hist(estimated_returns[,3], breaks = 100)
hist(estimated_returns[,4], breaks = 100)
par(mfrow = c(1, 1))

original_Lstocks <- day_value[-1] %>% as.numeric %>% log #logged values of final closing price before date

## adds logged final price to estimated returns
changed_vals <- t(apply (estimated_returns, 1, function(x) x + original_Lstocks))

estimated_stocks <- exp(changed_vals)
head(estimated_stocks)

## These are our estimates for the potential stock price
par(mfrow = c(4,1))
hist(estimated_stocks[,1], breaks = 100)
hist(estimated_stocks[,2], breaks = 100)
hist(estimated_stocks[,3], breaks = 100)
hist(estimated_stocks[,4], breaks = 100)
par(mfrow = c(1,1))



## estimate the portfolio value changes
dim(estimated_stocks)

estimated_value <- estimated_stocks %*% as.numeric(my_shares[-1])
current_value <- as.numeric(day_value[-1]) %*% as.numeric(my_shares[-1])

estimated_losses <- apply (estimated_value, 1, function(x) x - current_value)
head(estimated_losses)
hist(estimated_losses, breaks = 100)

hist(estimated_losses, breaks = 100, xlim = c(-8000, 6000))
ggplot(as_tibble(estimated_losses), aes(value)) + geom_histogram()

sorted_ELosses <- sort(estimated_losses)
copula_VaR <- quantile (sorted_ELosses, probs = 1 - alpha, type = 7)
copula_AVaR <- mean(estimated_losses[estimated_losses <= copula_VaR])

sprintf("The copula estimated VaR is %f, and the copula estimated AVaR is %f at the %.2f confidence interval",
        copula_VaR, copula_AVaR, 1-alpha)

sprintf("The linear model estimated VaR is %f, and the linear model estimated AVaR is %f at the %.2f confidence interval",
        lin_VaR, lin_AVaR, 1-alpha)

head(estimated_value)
mean(estimated_value)
hist(estimated_value)

day_value <- close_prices %>% dplyr::filter (Date == end_date_1)


as.numeric(day_value[-1]) %*% as.numeric(my_shares[-1])
fit_1


```
